# Reproducible Research: Peer Assessment 1

### Loading and preprocessing the data

Unzipping and loading data. 
```{r data-loading}
unzip("activity.zip")
activity <- read.csv(file="activity.csv")
```

Transforming the date into POSIXlt format.

```{r dataformatting}
activity$date <- strptime(activity$date, format="%Y-%m-%d")
```

### What is mean total number of steps taken per day?

Computing a histogram of the total number of steps taken each day using plyr package.

```{r sumsteps}
library(plyr)
StepsSummary <- ddply(activity, .(date), summarize, 
      sumSteps = sum(steps))

hist(StepsSummary$sumSteps, col = "grey", breaks = 10, main = c("Histogram of Number of Steps"), 
xlab = c("Number of Steps per Day"))
```

```{r meanmedian}
Mean = as.numeric(round(mean(StepsSummary$sumSteps, na.rm = TRUE), 0))
Median = median(StepsSummary$sumSteps, na.rm = TRUE)
```

For the total number of steps taken each day, the mean is **`r Mean`** and the median is **`r Median`**.

### What is the average daily activity pattern?

Making a time series plot to examine the daily activity pattern. of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r sumdailysteps}
require(plyr)
StepsTimeInt <- ddply(activity, .(interval), summarize, 
      timeIntSteps = round(mean(steps, na.rm = TRUE), 0))

library("ggplot2")
ggplot(StepsTimeInt, aes(x = interval, y = timeIntSteps)) + geom_line() 

maxRow <- which.max(StepsTimeInt$timeIntSteps)
tInt <- StepsTimeInt$interval[maxRow]
```

Of the 5-minute intervals, on average across all the days in the dataset, the interval **`r tInt`** (8:35am) had the maximum number of steps.


### Imputing missing values

Imputing missing values (`NA`)to avoid introducing bias into calculations or summaries of the data. Calculating the number of missing values first.

```{r NAvalues}
NAvalues = sum(colSums(is.na(activity)))
```

Total number of missing values in the dataset is **`r NAvalues`**.

Filling in all of the missing values in the dataset based on the the mean for that 5-minute interval.
Steps are as follows:
- Creating an index of missing values
- Creating a vector of step mean that correspons to each given interval for each day based on the previous calculation of step means
- Inserting values for missing observations, while populating remaining rows with NAs
- Replacing all NAs with 0 and summing up steps and average steps columns to create imputed values.

```{r imputingNAValues}
NAindex <- !complete.cases(activity$steps)

stepsMean = StepsTimeInt$timeIntSteps[which(StepsTimeInt$interval == activity$interval)]
activity$stepsMean[NAindex] <- stepsMean[NAindex]
activity[is.na(activity)] <- 0
activity$stepsNoNAs <- activity$steps + activity$stepsMean

```

Creating a new dataset with the missing data filled in.

```{r imputeddataset}
activityImputed <- subset(activity, select = c(stepsNoNAs, date, interval))
```

Making a histogram of the total number of steps taken each day.

```{r newhistogram}
StepsSummaryImputed <- ddply(activityImputed, .(date), summarize, 
      sumSteps = sum(stepsNoNAs))

hist(StepsSummaryImputed$sumSteps, col = "grey", breaks = 10, main = c("Histogram of Number of Steps With Imputed Missing Values"), 
xlab = c("Number of Steps Taken per Day"))
```

Calculate and report the **mean** and **median** total number of steps taken per day. 

```{r meanmedianimputed}
MeanImputed = round(mean(StepsSummaryImputed$sumSteps, na.rm = TRUE), 0)
MedianImputed = median(StepsSummaryImputed$sumSteps, na.rm = TRUE)
```

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Mean and median values are lower with imputed values -- since the distribution has a right-skew, imputed values on average are smaller. The difference between the Mean and the Median increases.

### Are there differences in activity patterns between weekdays and weekends?

Using the dataset with the filled-in missing values, created another column with a weekday label using weekeays() function. Adjusted the levels to only include "weekday" and "weekend".

```{r weekdaypatterns, fig.width = 8}
activityImputed$weekdays <- as.factor(weekdays(activityImputed$date))
labelWeekdays = c("weekday", "weekday", "weekend", "weekend", "weekday", "weekday", "weekday")
levels(activityImputed$weekdays) = as.factor(labelWeekdays)
```

Creating a panel plot for the time series with the 5-minute interval as x-axis and the average number of steps taken, averaged across all weekday days or weekend days as y-axis. The plot illustrates differences in activity between weekend and weekdays.

```{r plotbyweekday}
require(plyr)
StepsTimeImputed <- ddply(activityImputed, .(interval, weekdays), summarize, 
      timeIntSteps = round(mean(stepsNoNAs),0))

library("ggplot2")
ggplot(StepsTimeImputed, aes(group = factor(weekdays), x = interval, y = timeIntSteps, color = weekdays)) + geom_line() 
```